---
title: "Section 1"
author: "Nadia Kennar, Research Associate with UKDS"
date: '2022-04-12'
output: html_document
---

This section will cover the following 

1. Time Series Data Representations 
2. Converting Time Series Object 
3. Making Decomposition Plots 
4. Checking for Stationarity 
5. Applying Rolling Averages


## Load Packages 

```{r}
library(crimedata)
library(dplyr)
library(fable)
library(tsibble)
library(ggplot2)
library(lubridate)
library(forecast)
library(TTR)
library(tseries)
library(readr)
library(ggfortify)
```



## Time Series Data Representation 


We will first start this code demo by exploring some different types of time series data 

Typically when working with data in R, you need to decide the object class of your data at hand. This is important because the object class you choice affects more than the data is stored; it will dictate which functions will be available for data pre-processing, analysing and plotting. 

Typically data in R is stored as a vector. However, when working with time series data we tend to have to convert to the right object class known as a 'time series object' 

R has at least eight different implementations of data structures for representing time series data. The list below identifies the most frequently used packages

> ts #base package 
> zoo 
> xts 
> tsibble 




## King - (non-seasonal)

We will first use a data set named 'Kings' which is an example of a small time series data. It records the age of death for 42 successive king of Enlgand.

The data set can be found here: <https://robjhyndman.com/tsdldata/misc/kings.dat>

AS this data is a text file we can read in the data using the *scan()* function. 


```{r}

king <- scan('https://robjhyndman.com/tsdldata/misc/kings.dat', skip = 3) 
king

class(king)
```



To convert a dataframe to a time series object we can use the 'ts' function from the base package 


```{r}
king <- ts(king)

class(king)
king
```





## Births - Seasonal Data 

What happens if we have data that has been collected at more regular intervals i.e. monthly, weekly, quarterly etc. 

If this is the case you would have to specify the number of times that day was collected pet year using the *frequency* parameter. 

Lets explore this with a different dataset, named 'Births' which refers to the number of births per month in New York City, from 1946 to 1958. 

The dataset can be found here: <"http://robjhyndman.com/tsdldata/data/nybirths.dat">, This time we can call on some more parameters, the code to do so would look like this *ts(vector, start=, end=, frequency=)*


```{r}
births <- scan("http://robjhyndman.com/tsdldata/data/nybirths.dat")
births


births <- ts(births, frequency = 12, start = c(1946, 1))
births

```






## Plotting a Time Series Dataset 

To plot both datasets (seaonsol or non-seasonal) you can use the
*plot.ts* function from the ts package. 



```{r}
plot.ts(king)
plot.ts(births)
```



## Plotting TS objects using Ggplot

```{r}
library(ggfortify)
autoplot(births)


autoplot(births, ts.geom = "ribbon", fill = "grey")

```



## Plotting with the Forecast package 

S/ARIMA = 
Seasonal (S) / Autoregressive (AR) / Integrated (I) / Moving Average (MA)


```{r}
library(forecast)

#run the model 
births_arima <- auto.arima(births, seasonal = T)
print(summary(births_arima))
checkresiduals(births_arima)

#make the forecast 
births_forecast <- forecast(births_arima, h = 12)
print(births_forecast)

#plot 
autoplot(births_forecast)

```






## Decomposition

*ts = seasonality (s) + trend (t) + noise (n) + cycle (c)

Additive compononent = [s+t+n*c]
Multiplicative Component = [s*t*n*c]


```{r births}
birth_decomp <- decompose(births)
head(birth_decomp)
plot(birth_decomp)

# you can improve the graphics by using the autoplot function from the forecast package
births %>% decompose %>% 
  autoplot()
```



## How to remove seasonality from your data (seasonal adjustment)?

*To seasonally adjust a additive model (Time Series - Seasonal)
*To seasonally adjust a multiplicative model (Time Series / Seasonal)

```{r}
adjusted_births <- births - birth_decomp$seasonal
plot(adjusted_births)
```






## Checking for Stationarity 

To reiterate,a time series is said weakly stationary if 

-  its mean is constant
-  its standard deviation is constant 
-  its cross covariance does not depend on time but depends only on the lag between the two series concerned. 

There are two ways to check for this; 


1. the graphical way 

```{r}
ts.plot(king)

```




2. the statistical method

```{r}
library(tseries)
adf.test(kings)
```


we can conduct is the Augmented Dickey–Fuller (ADF) t-statistic test to find if the series has a unit root (a series with a trend line will have a unit root and result in a large p-value).






## The Zoo Package - how to calculate basic rolling values 


```{r}
install.packages("zoo")
require("zoo")

#load the avergae temperatire by month and year in Nottingham
nottem

#plot 
plot(nottem)

#smooth 
not_mean <- rollmean(nottem, 12,fill = NA, align = "right")
not_mean
plot(not_mean)

```





## The XTS package 

XTS stands for eXtensible time series. It is an extension of the zoo object that includes a matrix and index. Xts is a matrix with associated times for each observation


```{r}
library(xts)

#creating xts object
data <- rnorm(10)
dates <- seq(as.Date("2016-01-01"), length = 10, by = "days")
xts <- xts(data, order.by = dates)
xts


#converting to a xts pbject 
xts2 <- as.xts(nottem)
xts2


```












